### 2019-12-17 latex_user_manual/chapters/introduction.tex
Initial commit.

% !TeX spellcheck = en_GB
\documentclass[review]{elsarticle}

\usepackage{textcomp, gensymb}
\usepackage{color}
\usepackage[table]{xcolor}
\usepackage{epstopdf}
\usepackage{subfig}
\usepackage[T1]{fontenc}
\usepackage{lineno}
\usepackage{amsmath}
\usepackage{hyperref}

\epstopdfsetup{outdir=./figures/}

\definecolor{gray090}{gray}{0.90}

\modulolinenumbers[5]

\journal{Environmental Modelling \& Software}

\bibliographystyle{elsarticle-num}

\begin{document}

	\begin{frontmatter}

		\title{SPAMDA: Software for Pre-processing and Analysis of Meteorological DAta to build datasets}

		\author[UCO]{Antonio M. G\'omez-Orellana}
		\ead{am.gomez@uco.es}

		\author[UCO]{Juan C. Fern\'andez\corref{JCFC}}
		\cortext[JCFC]{Corresponding author.}
		\ead{jfcaballero@uco.es}

		\author[UCO]{Manuel Dorado-Moreno}
		\ead{manuel.dorado@uco.es}

		\author[UCO]{Pedro A. Guti\'errez}
		\ead{pagutierrez@uco.es}

		\author[UCO]{C\'esar Herv\'as-Mart\'inez}
		\ead{chervas@uco.es}

		\address[UCO]{Department of Computer Science and Numerical Analysis, University of Cordoba, C\'ordoba, Spain.}

		\begin{abstract}
			%Meteorological data play an important role for the comprehension of the environment, and they are extensively used to perform environmental learning. Machine Learning (ML) techniques have become a valuable support in many research areas, improving the performance of traditional statistical procedures. However, such techniques require datasets containing information related to the topic under study, which are not always available in an appropriate format. Preparing these datasets implies a lot of time and effort by the researchers. This paper presents a software tool for creating datasets using meteorological observations from two well-known sources of information: the \textit{National Data Buoy Center} (NDBC) and the \textit{National Centers for Environmental Prediction} (NCEP)/\textit{National Center for Atmospheric Research} (NCAR) \textit{Reanalysis Project}. The datasets created by the software are ready to be used as inputs for ML techniques in prediction tasks (classification or regression). As the designed software is able to simplify the creation of the datasets and reduce the time needed for this task, it prevents researchers from performing repetitive technical work, allowing them to concentrate on the study of the meteorological aspects of the data. Therefore, researchers can benefit from it in order to achieve a more efficient exploitation and protection of the environment.
			Meteorological data play an important role for the comprehension of the environment, and they are extensively used to perform environmental learning. Machine Learning (ML) techniques, a valuable support in many research areas, require datasets containing information related to the topic under study, which are not always available in an appropriate format, and its preparation implies a lot of time and effort by the researchers. This paper presents a software for creating datasets using meteorological observations from two well-known sources of information: the \textit{National Data Buoy Center} and the \textit{National Centers for Environmental Prediction and for Atmospheric Research Reanalysis Project}. As the designed software is able to simplify the creation of the datasets (ready to be used as inputs for ML techniques in prediction tasks) and reduce the time needed for this task, researchers can benefit from it in order to achieve a more efficient exploitation and protection of the environment.
		\end{abstract}

		\begin{keyword}
			Meteorological data\sep Reanalysis data\sep  Pre-processing\sep Creating datasets\sep  Prediction tasks\sep Marine energy
		\end{keyword}

	\end{frontmatter}

	\linenumbers

	\section{Introduction}

		A better understanding of the environment is of vital importance for science, contributing not only to more efficient exploitation of natural resources but also to the development of new strategies aimed at its protection. In that sense, meteorological observations provide an essential and valuable source of information which is widely used by researchers to address environmental learning, comprehension, prediction and conservation in numerous oceanic and atmospheric studies of a wide variety of areas (e.g. climate change, agriculture, energy, etc.). Some specific examples of the diversity of fields in which meteorological data can be used in are, among others: directional analysis of sea storms \cite{LAFACE201545}, wind power ramp events prediction \cite{DORADOMORENO2017428}, offshore wind energy assessment \cite{DVORAK20101244}, study of the responses exhibited by plankton to fluid motions \cite{FUCHS2016109}, trends in solar radiation \cite{SILVA20101852} or simulation of extreme near shore sea conditions \cite{GOULDBY201415}. All these studies require a prior data collection and its adaptation to a specific format that allows the interpretation of them.

		On the other hand, special purpose software is usually developed to help researchers to advance in their studies related to energy and environmental modelling, becoming a great support for decision-making in the exploitation and protection of the environment. In \cite{CAMARGONOGUEIRA2016361}, a software tool for designing solar water heating systems is developed, which simulates different situations and finds the best technical and economical solution. In \cite{LO2015293}, an integrated simulation tool for the optimum design of bifacial solar panel with reflectors is presented. This tool can also be used to analyse the performance of the solar cells. A framework for data integration of offshore wind farms is implemented in \cite{NGUYEN2013150} in order to facilitate data exchange and improve operation and maintenance practices. In \cite{GUTIERREZ201369}, a tool based on \textit{Fatigue, Aerodynamics, Structures, and Turbulence} (FAST) code for performing design optimisation of offshore wind turbines is presented, which can analyse a massive group of cases by means of its parametric design capability. A design tool for architects based on \textit{Lighting and Thermal} (LT) method is used in \cite{BAKER2013156}, with the purpose of comparing and optimising design solutions in terms of energy and comfort performance. Raabe et al. \cite{RAABE2010213} developed two software tools, \textit{Model of Equilibrium of Bay Beaches} (MEPBAY) and \textit{Coastal Modeling System} (SMC), to support different operational levels of headland-bay beach in coastal engineering projects, and an interactive web application, called the \textit{Stream Hydrology And Rainfall Knowledge System} (SHARKS), for analysis of hydrologic and meteorological time series data that can be used for providing valuable insights to urban hydrologic processes is presented in \cite{BRENDEL2019}.

		Marine energy prediction is currently a hot topic where meteorological data is used in. Marine Renewable Energy (MRE) is one of the most important renewable and sustainable energy sources available in our environment, and it includes ocean thermal energy, marine tidal current energy and wave energy, among others. Its benefits and great potential \cite{ZEYRINGER20181281} make it one of the most relevant natural resources, playing a crucial role not only in the reduction of the emission of greenhouse gases but also in all other aspects involved in the difficult challenge of the transition to a low carbon footprint society \cite{BHATTACHARYA2017157, BREDE201344}. Wave energy exhibits a more stable power supply than wind energy and even solar energy. In recent years \textit{Wave Energy Converters} (WECs) \cite{FALCAO2010899} have been developed to transform this wave energy into electricity. WECs are mechanical devices that convert kinetic energy into electrical energy by means of either the vertical oscillation of waves or the linear motion of them. Nevertheless, waves are difficult to be characterised due to their stochastic nature, because of the influence of a large number of environmental factors that exert on them \cite{ochi1998}. As a consequence of this complexity, many aspects of WEC design, deployment and operation \cite{CROWLEY2018159, Abdelkhalik2016, 6898109} need a proper prediction of waves, in order to maximise the wave energy extraction. For this purpose WECs use wave \textit{flux of energy} ($F_e$) which can be calculated from the two most important wave parameters in this regard: \textit{significant wave height} ($H_s$) and \textit{wave energy period} ($T_e$). Additionally, wave predictions are also helpful for designing offshore structures \cite{CHATZIIOANNOU2017126}, operational works in the sea \cite{DALGIC2015211}, providing technical information to the coastal and coral reef planners in developing countries \cite{CALLAGHAN2018123}, etc.

		Currently, and as a support to traditional study procedures, Machine Learning (ML) techniques \cite{Alpaydin:2004:IML:1036287,Bishop:2006:PRM:1162264} are being widely used in numerous research fields related to classification, regression and optimisation tasks \cite{srivastava2017large}, obtaining significant improvements in the performance of the results. ML methodologies can be used not only by experienced data and computer scientists but also by other researchers. For example, the well-known \textit{Waikato Environment for Knowledge Analysis} (WEKA) \cite{WEKA} software tool provides researchers with a wide collection of ML algorithms. In this way, ML techniques have been already applied to tackle wave characterisation, accurately estimating $H_{s}$ and $T_{e}$ parameters \cite{DURANROSAL2017268, KUMAR2017605}, given that robustness of ML methods can tackle the previously explained difficulties in wave energy prediction. The problem is that, in order to apply these methods, it is essential to obtain datasets with relevant information about the issue under study, used to infer knowledge. Usually, these datasets are not publicly available in a friendly format, and their generation is the first step needed.

		The information to create these datasets can be obtained from meteorological observations, but such information may be available in a inappropriate format and even contain missing values or measurements. Consequently, it is usually required to perform pre-processing tasks for improving the quality of the data, such as the replacement of missing values, outlier detection or data normalisation, among others. Furthermore, if more than one source of information is used to achieve a better characterisation of the problem under study \cite{JOHANSSON2015143, FERNANDEZ201544, Adams2010}, then a merging or matching process have to be carried out by the researchers to manually create the datasets with the needed information. Moreover, depending on the subject and the ML technique to be applied, or even if the researcher considers other factors in order to improve the results or have more in-depth conclusions, the datasets would have to be updated afterwards. In summary, many important details and different intermediate steps have to be considered when creating suitable datasets for ML techniques, resulting in an extremely tedious task.

		The main purpose of this paper is to present a software tool able to prevent researchers from performing this repetitive work and greatly simplify all the steps involved in the creation of datasets. The meteorological data considered for the tool come from two well-known sources of information: the \textit{National Oceanic and Atmospheric Administration} (NOAA) \textit{National Data Buoy Center} (NDBC) \cite{NOAA} and the \textit{National Centers for Environmental Prediction} (NCEP)/\textit{National Center for Atmospheric Research} (NCAR) \textit{Reanalysis Project} (NNRP or R1) \cite{Kalnay1996, Kistler2001}. The software tool presented in this work is named SPAMDA (Software for Pre-processing and Analysis of Meteorological DAta to build datasets). As SPAMDA performs all this data processing, it reduces the time involving these tasks and allows the researchers focus on the study of the meteorological aspects of the observations. The datasets obtained are ready to be used as input for ML techniques in prediction tasks (classification or regression), although the researchers can use them for other purposes. These datasets contain one or more meteorological variables as inputs and one variable as target (variable to be predicted). The format of the generated datasets will be \textit{Attribute-Relation File Format} (ARFF) \cite{WEKA_ARFF}, which is the one used by WEKA. Besides, the datasets can also be generated in \textit{Comma-Separated Values} (CSV) format, enabling the researchers to use others tools.

		Up to now, and to the best of our knowledge, this is the first software tool addressing the problem previously discussed, combining meteorological data from NDBC and NNRP. The advantages that SPAMDA offers to the researchers will be detailed in Section \ref{sec:SPAMDA}, although some of them are briefly summarised below:
		\begin{itemize}
			\item The generation of datasets becomes a very easy and customizable task, by means of the selection of different input parameters.
			\item It makes the researcher focus on oceanic and atmospheric studies, without having to worry about mechanical tasks.
			\item It provides information about the quality and quantity of the data.
			\item It avoids possible researcher errors in the intermediate steps of the process of creation of the datasets.
			\item It includes different pre-processing tasks, such as normalisation and missing data recovery.
			\item It facilitates data management and well-organised storage of the datasets.
			\item Its modular design allows the implementation of new functional modules for managing meteorological data from others sources for renewable energy research.
			\item It includes an user-friendly GUI, facilitating and greatly simplifying data management, and it is integrated with the Explorer environment of WEKA.
			\item It is multi-platform, and it can be used on any computer with Java regardless of the operating system.
		\end{itemize}

		This paper is organised as follows: Section \ref{sec:DataSources} describes the sources of information used by SPAMDA for creating datasets. Section \ref{sec:SPAMDA} describes in detail the features of the software tool. Section \ref{sec:CaseStudy} shows a case study describing the use of SPAMDA in a practical approach. Section \ref{sec:Conclusions} provides the final conclusions and future work.

@@

		The data provided by the above-mentioned sources of information of SPAMDA is described below:

@@

			\item NDBC is a part of the \textit{National Weather Service} (NWS). NDBC designs, develops, operates, and maintains a network of data collecting buoys (stations). The mission of the network is to collect real-time marine meteorological and oceanographic observations, such as $H_s$, dominant wave period, or wind speed and direction, among others.

			The buoys maintained by NDBC are deployed in the coastal and offshore waters around oceans and seas, and they are equipped with assorted sensors which allow them to perform different measurements. The information collected by the buoys is available on the NDBC website \cite{NOAA_1}, and it is divided into different groups. One of them corresponds to standard meteorological information of the historical data collected by each buoy, which can be downloaded as annual text files and whose format was adopted by NDBC since January 2007 \cite {NOAA_2}. These files contain hourly measurements per day from $00$:$50$ to $23$:$50$ UTC (Universal Time Coordinated) and from $23$:$50$ 31th December of the previous desired year to $22$:$50$ 31th December of the desired year. In Table \ref{tab:measurementsDescription}, a comprehensive measurement description and the corresponding units are provided as a summary for the reader. A fragment of one of these files, which contains the measurements collected during year $2017$ by the buoy identified as \textit{Station 46001} in NDBC, is shown in Fig. \ref{fig:fragmentAnnualTexFile}. Each column corresponds to a meteorological variable or attribute, and each row or instance corresponds to the values of the measurements collected by the buoy for each attribute at a specific date and time.

			\begin{figure}[ht!]
				\centering
				\includegraphics[scale=0.50]{figures/FigureFragmentAnnualTextFile.png}
				\caption{A fragment of an annual text file of the \textit{Station 46001}.}
				\label{fig:fragmentAnnualTexFile}
			\end{figure}

			\begin{table}[!ht]

@@

				\begin{tabular}{ccm{8.60cm}@{\setlength{\tabcolsep}{0pt}}m{0.0cm}}

@@

@@

					WDIR & degT & Wind direction (the direction the wind is coming from in degrees clockwise from true North) during the same period used for WSPD. \\

@@

					ATMP & degC & Air temperature (Celsius degrees). &\\[0.10cm]

					\cellcolor{gray090}WTMP & \cellcolor{gray090} degC & \cellcolor{gray090} Sea surface temperature (Celsius degrees). For buoys the depth is referenced to the hull's waterline. For fixed platforms it varies with tide, but is referenced to, or near Mean Lower Low Water (MLLW).\\

@@

					TIDE & ft & The water level in feet above or below MLLW. \\

@@

			Note that the data collected by the network of buoys may be incomplete due to diverse circumstances such as the weather conditions in which the buoys have to operate, failures or malfunctioning elements of the buoys, among others. Accordingly, it may be the situation that some of the measurements are completely missing (missing date or instance) or partially missing (some measurements not recorded), by a buoy or by a set of buoys, once in a while or over a period of time. It may be also possible that the measurements have been recorded at a time different from the expected one. This aspect have to be taken into account when creating the datasets. This casuistry is explained in detail in Appendix A.

			\item NNRP provides three-dimensional global reanalysis of numerous meteorological variables (e.g. air temperature, components South-North and West-East of wind speed, relative humidity, pressure, etc.), which is available monthly, daily and every $6$ hours at $00$ Z (Zulu time), $06$ Z, $12$ Z and $18$ Z from $1948$ on a global $2.5\degree$ x $2.5\degree$ grid. Weather observations are from different sources, such as ships, satellites and radar, among others. Reanalysis data is created assimilating such observations using the same climate model throughout the entire reanalysis period in order to reduce the effects of modelling changes on climate statistics. Such information has become a substantial support of the needs of the research community, even more in locations where instrumental (real time) data is not available.

			The reanalysis data is available in the NNRP website \cite{NNRP}, which it is accessible through different sections. Such data can be fully (a global $2.5\degree$ x $2.5\degree$ grid) or partially (only the desired reanalysis nodes or sub-grid) downloaded as \textit{Network Common Data Form} (NetCDF) files \cite{NetCDF}, a special binary format for representing scientific data, which provides a description of the file contents and also includes the spatial and temporal properties of the data. Each reanalysis file contains the values of a meteorological variable estimated by a mathematical model for each reanalysis node. For a better understanding, in Fig. \ref{fig:subGrid} an approximate representation of a sub-grid containing six reanalysis nodes around the geographical location of a buoy (obtained from NDBC) is shown.

@@

		Therefore, with both sources of information, which complement each other, and carrying out a matching process, SPAMDA will create datasets for prediction tasks. In this way, the dataset input variables will be one or more reanalysis variables from NNRP and one or more measurements from NDBC. The dataset output variable will always be one measurement from NDBC.

	\section{SPAMDA}\label{sec:SPAMDA}

		SPAMDA combines  meteorological information from NDBC and NNRP to obtain new datasets for oceanic and atmospheric studies. In order to do so, SPAMDA manages three different types of datasets, which will be described in detail in the following sections, but are briefly introduced bellow for giving the reader a better general understanding:
			\begin{itemize}
				\item \textit{Intermediate datasets}: They contain the meteorological observations from NDBC.
				\item \textit{Pre-processed datasets}: They are obtained as a result of pre-processing tasks performed on the intermediate datasets.
				\item \textit{Final datasets}: Created by merging an intermediate or pre-processed dataset (which contain the information from NDBC) with the reanalysis data from NNRP. This procedure is referenced in SPAMDA as matching process and will be carried out according to the study to be performed (classification or regression).
			\end{itemize}

		\begin{figure}[ht!]
			\centering
			\includegraphics[scale=0.32]{figures/FigureSPAMDA.eps}
			\caption{Brief outline of the functionality provided by SPAMDA.}
			\label{fig:SPAMDA}
		\end{figure}
		SPAMDA consists of three main functional modules, whose main features, represented in Fig. \ref{fig:SPAMDA}, are the following:
		\begin{itemize}

			\item \textit{Manage buoys data}: The aim of this module is to provide features for the management and analysis of the information related to the buoys from NDBC. This includes:
			\begin{enumerate}
				\item Entering and updating the information of each buoy.
				\item Creation of intermediate datasets with the collected measurements.
				\item Pre-processing tasks for obtaining the pre-processed datasets.
				\item Matching process to merge the information from NDBC and NNRP
				\item Creation of the final datasets accordingly to the ML technique to use (classification or regression).
			\end{enumerate}

			\item \textit{Manage reanalysis data}: This module is used for the management of the reanalysis data provided by the NNRP. In this way, the researchers can keep the reanalysis data files updated for their studies. Such files will be used, depending on the researchers needs, in the matching process when obtaining the final datasets.

			\item \textit{Tools}: This module includes features for converting intermediate or pre-processed datasets to ARFF or CSV format and for opening ARFF files with WEKA software.

		In the following subsections each integrated functional module is described in detail.

			\subsection{Buoys}\label{sec:Buoys}

				When a new buoy is included in SPAMDA the following information, which can be obtained from NDBC, is requested:
				\begin{itemize}
					\item \textit{Station ID}: An alphanumeric identifier that allows easy identification of the buoy.
					\item \textit{Description}: A short description of the buoy.
					\item \textit{Latitude}: North or South geographical localisation (degrees) of the buoy.
					\item \textit{Longitude}: West or East geographical localisation (degrees) of the buoy.
					\item \textit{Measurements files}: The above-mentioned annual text files of the standard meteorological information collected by the buoy and downloaded from the NDBC website. This will be used for the creation of the intermediate datasets. One file per year is expected.
				\end{itemize}

				For clarification, an example is presented in Fig. \ref{fig:buoys}, where the buoy ID1 has three annual text files and the buoy ID2 has two annual text files.

				\begin{figure}[ht!]
					\centering
					\includegraphics[scale=0.42]{figures/FigureBuoys.eps}
					\caption{Example of entering two buoys with its annual text files.}
					\label{fig:buoys}
				\end{figure}

			\subsection{Datasets}\label{sec:Datasets}

				Once a buoy has been included as described in Section \ref{sec:Buoys}, it is possible to create datasets with one or more annual text files, which are referenced in SPAMDA as intermediate datasets. In this module, the researchers can manage intermediate datasets of each buoy, which are the baseline for their studies, by creating new ones or deleting the unnecessary ones.

				When an intermediate dataset is created, it is associated with its corresponding buoy. Besides, a summary of its content is also created, providing relevant information such as the number of instances, the dates of the first and last measurements, the annual text files included, and the missing and duplicated dates.

				An example where three intermediate datasets have been created is presented in Fig. \ref{fig:datasets}. The two intermediate datasets of the buoy ID1 contain meteorological data of different years, and the intermediate dataset of the buoy ID2 contains meteorological data of two years. For each buoy, as many intermediate datasets as needed can be created.

				\begin{figure}[ht!]
					\centering
					\includegraphics[scale=0.40]{figures/FigureDatasets.eps}
					\caption{Example of the creation of the intermediate datasets.}
					\label{fig:datasets}
				\end{figure}

			\subsection{Pre-process} \label{sec:Preprocess}

				Data pre-processing prepares the raw data (intermediate datasets) to be able to be treated correctly by ML algorithms. In this way, the quality of dat can be improved prior to the learning phase, by applying pre-processing tasks (filters). The result will be referenced as pre-processed datasets.

				SPAMDA provides several filters grouped in three categories, \textit{Attribute}, \textit{Instance} and \textit{Recover missing data}, including the configuration of their parameters and a short description of them:

				 \item \textit{Attribute}: All theses filters can be applied to the attributes (variables of the buoy from NDBC) of the intermediate dataset.

					\begin{itemize}
						\item \textit{Normalize} \cite{WEKA_Filter_Normalize}: This filter normalises all numeric values of each attribute. The resulting values are by default in the interval [0,1].
						\item \textit{Remove} \cite{WEKA_Filter_Remove}: It removes an attribute or a range of them.
						\item \textit{RemoveByName} \cite{WEKA_Filter_RemoveByName}: It removes attributes based on a regular expression matched against their names.
						\item \textit{ReplaceMissingValues} \cite{WEKA_Filter_ReplaceMissingValues}: For each attribute, all the missing values will be replaced by the average value of the attribute.
						\item \textit{ReplaceMissingWithUserConstant} \cite{WEKA_Filter_ReplaceMissingWithUserConstant}: This filter replaces all the missing values of the attributes with an user-supplied constant value.
					\end{itemize}

				 \item \textit{Instance}: All theses filters can be applied to the instances (hourly measurements of the buoy from NDBC) of the intermediate dataset.
					\begin{itemize}
						\item \textit{RemoveDuplicates} \cite{WEKA_Filter_RemoveDuplicates}: With this filter, all duplicated instances are removed.
						\item \textit{RemoveWithValues} \cite{WEKA_Filter_RemoveWithValues}: This filter removes all the instances that match the attribute and the value supplied by the user.
						\item \textit{SubsetByExpression} \cite{WEKA_Filter_SubsetByExpression}: It removes all the instances which do not match a user-specified expression.
					\end{itemize}

				 \item \textit{Recover missing data}: All these filters can be applied to the instances of the intermediate dataset.
					\begin{itemize}
						\item \textit{Replace missing values with next nearest hour}: The missing values of each attribute are replaced with the next nearest non missing value.
						\item \textit{Replace missing values with previous nearest hour}: This filter replaces the missing values of each attribute with the previous nearest non missing value.
						\item \textit{Replace missing values with next $n$ hours mean}: The missing values of each attribute are replaced with the next $n$ nearest non missing values mean, where $n$ can be configured by the user.
						\item \textit{Replace missing values with previous $n$ hours mean}: This filter replaces the missing values of each attribute in the intermediate dataset with the previous $n$ nearest non missing values mean.
						\item \textit{Replace missing values with symmetric $n$ hours mean}: The missing values of each attribute in the intermediate dataset are replaced with the $n$ previous and $n$ next non missing values mean.
					\end{itemize}

				SPAMDA allows the researchers to undo the last filter applied or to restore the initial content of the intermediate dataset. Besides, the content and relevant statistical information (number of instances with missing values, minimum and maximum values, mean and standard deviation) of the intermediate and the pre-processed datasets can be visualised in this module.

				Fig. \ref{fig:preprocess} shows an example where the intermediate datasets 1 and 2 of the buoy ID1 have been pre-processed, obtaining as a result the pre-processed dataset 1 of each one. The intermediate dataset 1 of the buoy ID2 has been also pre-processed. \textit{Pre-processed dataset n} represents that the researchers can create as many pre-processed datasets as they consider opportune.

				\begin{figure}[ht!]
					\centering
					\includegraphics[scale=0.36]{figures/FigurePreprocess.eps}
					\caption{Example of the creation of pre-processed datasets.}
					\label{fig:preprocess}
				\end{figure}

			\subsection{Matching configuration}
			\label{sec:matching_conf}

				In order to merge and format the data provided by the two sources of information described in Section \ref{sec:DataSources}, it is necessary to carry out a matching process. The matching procedure is performed using an intermediate or pre-processed dataset, which includes the measurements collected by a buoy from NDBC, and the needed reanalysis data files from NNRP. Note that SPAMDA is able to manage the NetCDF binary format for handling the information stored in the reanalysis files.

				Such process merges the information of both sources that match on time, but, given that the measurements of the buoys are hourly collected from $00$:$50$ to $23$:$50$ UTC, and the reanalysis data is available every $6$ hours at $00$ Z, $06$ Z, $12$ Z and $18$ Z, the matching can only be carried every $6$ hours (discarding the rest of measurements from the buoy data). Besides, and since there is still a difference of 10 minutes, the matching with the reanalysis data will be performed with the nearest buoy measurement (before or after) within a maximum of 60 minutes of difference. Finally, the matched instances of both sources will form the final datasets.

				Fig. \ref{fig:matchingProcess} presents an example of matching with the measurements collected during $2017$ by \textit{Station 46001} (NDBC) and the reanalysis data (NNRP) of the variable \textit{pressure} for reanalysis nodes $57.5$ N $\times$ $147.5$ W and $55.0$ N $\times$ $147.5$ W in the same year. In this way, only the instances from both sources that are linked with arrows (highlighted in green colour) will be used in the creation of the final datasets. Although the reanalysis dates have been presented in a human readable format, note that reanalysis dates are stored in hours from $01$-$01$-$1800$, and they have to be transformed for comparison taking into account the time zone. Such transformation is automatically done by SPAMDA when matching the instances.

				The reader can check in Appendix A \ref{app:AppendixA} an example with a more complex case of the procedure.
				\begin{figure}[ht!]
					\centering
					\includegraphics[scale=0.38]{figures/FigureMatchingProcess.png}
					\caption{An example of matching the data from NDBC (left) and NNRP (right).}
					\label{fig:matchingProcess}
				\end{figure}

				SPAMDA allows the researchers to perform a customisable matching process, for obtaining as many different versions of the same meteorological data as needed.
				Prediction tasks are based on the estimation of the output attribute using the information provided by the input attributes. Depending on the task, the datasets must be prepared and configured differently:				
				\begin{itemize}
					\item \textit{Classification}: The final datasets will be ready to use as input for ML classifiers, requiring a nominal output attribute, whose specific preparation is detailed in Section \ref{sec:FinalDatasets}.
					\item \textit{Regression}: The final datasets will be ready to use as input for regression methods, requiring a real output attribute, whose preparation is also explained in Section \ref{sec:FinalDatasets}.
					\item \textit{Direct matching}: In this case the inputs attributes have a direct correspondence with the output attribute, and it is not necessary to perform any additional preparation. Both input and target attributes are synchronised in time, in such a way that the final dataset is not intended for prediction purposes. For example, the final datasets may be used in lost data recovering tasks, in correlation studies, in descriptive analyses, etc.
				\end{itemize}

				The following parameters can be specified for the matching process:
				\begin{itemize}

					\item \textit{Flux of energy} \cite{FERNANDEZ201544}: When the $F_e$ is selected, it will be used as output. This attribute is not collected by the buoys, but it can be calculated from two wave parameters: $H_s$ and $T_e$, which are collected as WVHT and APD attributes, respectively, and were described in Table \ref{tab:measurementsDescription}. In this way, SPAMDA obtains the $F_e$ of each instance using the following equation:
					\begin{equation}
							F_e = 0.49 \cdot H^2_s \cdot T_e,
							\label{eq:fluxOfEnergy}
					\end{equation}

					where $F_e$ is measured in kilowatts per meter, $H_s$ is measured in meters and $T_e$ is measured in seconds. Note also that $F_e$ is defined in Eq. \ref{eq:fluxOfEnergy} as an average energy flux ($H_s$ is a kind of average wave height), though for simplicity it will be referred just as flux of energy.

					\item \textit{Attribute to predict}: Instead of using $F_e$, researchers can select any of the attributes collected by the buoys as output (e.g. significant wave height, WVHT, wind direction, WDIR, sea level pressure, PRES, etc.). Therefore, they can conduct different studies by selecting one attribute or other.

					\item \textit{Reanalysis data files}: In order to have a possible better description of the problem under study, more than one reanalysis variable can be considered as input. Remember that these files have to be previously downloaded from the NNRP website \cite{NNRP}, which should set the range of dates (temporal properties) and the desired sub-grid (spatial properties, see Fig. \ref{fig:subGrid}) for each variable of reanalysis.

					In that sense, the reanalysis data files must have the same spatial and temporal properties but related to different variables. SPAMDA simplifies this task by showing the reanalysis data files that are compatibles each other, and checking that the selection made by the researches meets that condition.

					\item \textit{Buoys attributes}: In addition to the reanalysis variables, the final datasets will also include the selected attributes as inputs (of the intermediate or pre-processed dataset used), providing a possible better characterisation of the problem under study, although it will depend on how correlated the attributes are.

					\item \textit{Include missing dates}: As above-mentioned, the information collected by a buoy may be incomplete due to measurements not recorded by it. As a consequence, the matching of instances between both sources of information may not be possible (missing dates). In that situation, the researchers can consider two options: 1) discard the instances affected or 2) include them. In the latter case, the final datasets will contain the affected instances, but the measurements of the buoy will be stored as missing values in WEKA format, denoted as \guillemotleft\textit{?}\guillemotright.

					\item \textit{Nearest reanalysis nodes to consider}: As already shown in Fig. \ref{fig:subGrid} (which represents six reanalysis nodes), the reanalysis data files may contain information of several reanalysis nodes. In this way, the researcher can:

							\item Consider all the reanalysis nodes contained in each file: in this case, the information provided by each reanalysis node contained in each selected reanalysis data file will be used.

>>>>

\chapter{Introduction to SPAMDA}

	\begin{onehalfspace}

		\section{SPAMDA description}

			SPAMDA is a software tool for creating datasets with meteorological data from two well-known sources of information, \textit{National Data Buoy Center} (NDBC) \cite{NOAA} and \textit{Reanalysis Project} (NNRP or R1) \cite{Kalnay1996, Kistler2001}.

			The datasets created with SPAMDA will be ready to use as input for Machine Learning (ML) techniques in classification or regression prediction tasks, although the researchers may use them in the way they deem suitable. These datasets will contain one or more meteorological variables as inputs and another one as target (variable to predict). The format of the datasets will be \textit{Attribute-Relation File Format} (ARFF) \cite{WEKA_ARFF} that it is used by the well-known tool \textit{Waikato Environment for Knowledge Analysis} (WEKA) \cite{WEKA}, which provides a wide collection of ML algorithms. Besides, the datasets can also be generated in \textit{Comma-Separated Values} (CSV) format, enabling the researchers to use others tools.

			Some of the advantages that SPAMDA tool offers are briefly summarised below:

				\begin{itemize}
					\item The generation of datasets becomes a very easy and customizable task, by means of the selection of different input parameters.
					\item It makes the researcher focus on oceanic and atmospheric studies, without having to worry about mechanical tasks.
					\item It provides information about the quality and quantity of the data.
					\item It avoids possible researcher errors in the intermediate steps of the process of creation of the datasets.
					\item It includes different pre-processing tasks, such as normalisation and missing data recovery.
					\item It facilitates data management and well-organised storage of the datasets.
					\item Its modular design allows the implementation of new functional modules for managing meteorological data from others sources for renewable energy research.
					\item It includes an user-friendly GUI, facilitating and greatly simplifying data management, and it is integrated with the Explorer environment of WEKA.
					\item It is multi-platform, and it can be used on any computer with Java regardless of the operating system.
				\end{itemize}

@@

		The data provided by the above-mentioned sources of information used by SPAMDA is briefly described below:

@@

			\item NDBC is a part of the \textit{National Weather Service} (NWS). NDBC designs, develops, operates, and maintains a network of data collecting buoys (stations). The mission of the network is to collect real-time marine meteorological and oceanographic observations, such as wave height, dominant wave period, or wind speed and direction, among others.

			The buoys maintained by NDBC are deployed in the coastal and offshore waters around oceans and seas, and are equipped with assorted sensors which allow them to perform different measurements. The information collected by the buoys is available in NDBC web page \cite{NOAA_1}, which is divided into different groups. One of them is the standard meteorological information of the historical data collected by each buoy, which can be downloaded as annual text files and whose format was adopted by NDBC since January 2007 \cite {NOAA_2}. These files contain hourly measurements per day from $00$:$50$ to $23$:$50$ UTC and from $23$:$50$ 31th Dec of the previous desired year to $22$:$50$ 31th Dec of the desired year. In Table \ref{tab:measurementsDescription} a comprehensive measurements descriptions and units of such information is provided.

			\clearpage

			\begin{table}[ht!]

@@

				\begin{tabular}{ccm{11.00cm}@{\setlength{\tabcolsep}{0pt}}m{0.0cm}}

@@

					%\textbf{Variable}&\textbf{Units}&\textbf{Description}&\\[0.30cm]

@@

					WDIR & degT & Wind direction (the direction the wind is coming from in degrees clockwise from true N) during the same period used for WSPD. \\

@@

					ATMP & degC & Air temperature (Celsius). &\\[0.10cm]

					\cellcolor{gray090}WTMP & \cellcolor{gray090} degC & \cellcolor{gray090} Sea surface temperature (Celsius). For buoys the depth is referenced to the hull's waterline. For fixed platforms it varies with tide, but is referenced to, or near Mean Lower Low Water (MLLW).\\

@@

					%PTDY & hPa & Pressure Tendency is the direction (plus or minus) and the amount of pressure change (hPa) for a three hour period ending at the time of observation. \\

					TIDE & ft & The water level in feet above or below Mean Lower Low Water (MLLW). \\

@@

			\item NNRP provides three-dimensional global reanalysis of numerous meteorological variables (e.g. air temperature, U/V-wind, relative humidity, pressure, etc.), which is available monthly, daily and every $6$ hours at $00$Z, $06$Z, $12$Z and $18$Z from $1948$ on a global $2.5\degree$ x $2.5\degree$ grid. Weather observations are from different sources, such as ships, satellites and radar, among others.

			%The reanalysis data is available in NNRP web page \cite{NNRP}, which are accessible through the different sections. Such data can be fully (a global $2.5\degree$ x $2.5\degree$ grid) or partially (only the desired reanalysis nodes or sub-grid) downloaded as \textit{Network Common Data Form} (NetCDF) files \cite{NetCDF}, a special binary format for representing scientific data which provides a description of what the data in each variable represents and the spatial and temporal properties of the data. Each reanalysis data file contains the estimated values by a mathematical model of one meteorological variable and for each reanalysis node. For a better understanding, in Fig. \ref{fig:subGrid} an approximate representation of a six sub-grid reanalysis nodes around the geographical localisation of a buoy (obtained from NDBC) is shown.

			The reanalysis data is available in NNRP web page \cite{NNRP}, which are accessible through the different sections. Such data can be fully (a global $2.5\degree$ x $2.5\degree$ grid) or partially (only the desired reanalysis nodes or sub-grid) downloaded as \textit{Network Common Data Form} (NetCDF) files \cite{NetCDF}, a special binary format for representing scientific data which provides a description of the file contents and also includes the spatial and temporal properties of the data. Each reanalysis file contains the values of a meteorological variable estimated by a mathematical model for each reanalysis node. For a better understanding, in Fig. \ref{fig:subGrid} an approximate representation of a sub-grid containing six reanalysis nodes around the geographical localisation of a buoy (obtained from NDBC) is shown.

@@

		With both sources of information SPAMDA will create datasets for prediction tasks. In this way, the input variables of the dataset will be one or more reanalysis variables from NNRP and one or more measurements from NDBC. The output variable of the dataset will be one measurement from NDBC. Note that the output variable cannot be used as input also.

	\end{onehalfspace}
